Interactive Visual Analysis of Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-09-18 draft.tex
Minor changes towards blind version

\author{\authorname{Paul Klemm\sup{1}, Sylvia Glaßer\sup{1}, Kai Lawonn\sup{1}, Marko Rak\sup{1} Henry Völzke\sup{2}, Katrin Hegenscheid\sup{3} and Bernhard Preim\sup{1}}
\affiliation{\sup{1} Department of Simulation and Graphics, University of Magdeburg, Germany}
\affiliation{\sup{2} Institute for Community Medicine, University of Greifswald, Germany}
\affiliation{\sup{3} Institute for Diagnostic Radiology and Neuro-Radiology, University of Greifswald, Germany}
\email{\{klemm, glasser\}@isg.cs.uni-magdeburg.de}
}

@@

% BLIND
%As described in a prior work (\cite{Klemm2014VIS}), the suggestion of potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
Suggesting potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
%%
We can achieve this also using the presented method by trying to describe several (dichotomous) target values with a decision tree constructed from all available data.
%%
This allows both to assess the discriminative power of the data set regarding the target variables, as well as the most important variables for the classification.
%Except for gender in the age-plot, no variable could be explained with reasonable accuracy for age or body size.
\paragraph{Predictive Power of Image-Derived Variables.}
We showed, that \emph{torsion}, \emph{curvature} and \emph{curvature angle} of the lumbar spine at the presented precision are not enough to characterize lumbar back pain in the SHIP data set.
%%
Our method made it possible to assess their discriminative power, which is largely limited for separating male and female subjects, \emph{nutrition} variables, as well as different disease indicators.
%%
The C4.5 algorithm showed to be an effective tool for evaluating a set of derived metrics regarding their suitability to classify non-image variables.
%%
Over-fitting to the data indicated by complex decision trees have to be taken into account as well.
%%
% \subsection{Proposed Features}
% \label{sec:ProposedFeatures}
% \begin{figure}[!h]
%   %\vspace{-0.2cm}
%   \centering
%   \includegraphics[width=0.475\textwidth]{figures/proposed-features}
%   \caption{
% 	Proposed features.
% 	%%
% 	(a) dented vertebraes are a sign of heavy stress and bone absorption.
% 	%%
% 	(b, c) the spine canal shape can be used to characterize scoliosis (curvature sagittal) and lordosis (curvature coronal) as well as the \emph{Cobb-Angles} \cite{Cobb}.
% 	%%
% 	(c) Spine canal thickness is associated with herniated disks.
% 	%%
% 	}
%   \label{fig:proposedFeatures}
% \end{figure}
% %%
% \noindent Pathological deformation of the lumbar spine is usually the last stage and is associated with very strong lower back pain.
% %%
% We have to capture earlier signs of pathological change in order to derive better predictive models.
% %%
%
% A very early sign of pathological deformation is the \emph{bone resorption in the center of a vertebra}.
% %%
% They change from being block-shaped to be dented in the center (Fig~\ref{fig:proposedFeatures}~a).
% %%
% This information can be obtained by segmenting the whole vertebra or the top and bottom point if each vertebra center.
% %%
% Another valuable variable would be the \emph{spine canal thickness} (Fig~\ref{fig:proposedFeatures}~c).
% %%
% Low spine canal thickness can be an indicator for an impending herniated disk.
% %%
% Both surface texture of the vertebra and thickness of the spine canal are used to diagnose herniated disks.
% %%
% The \emph{overall spine canal shape} is also of interest, since scoliosis and lordosis can be characterized more precise by deriving the Cobb angles from this shape (Fig~\ref{fig:proposedFeatures}~b,~c).
%% ------------------------------------
\paragraph{Future Work.}
\com{ToDo: Passen die Contributions hier wirklich?}
%% Contributions
We provide a method for comparative analysis of decision trees independent of the variable manifestation count using interactive scatter plots.
%%
We applied the technique to derive insight into the predictive power of 9 image-derived variables for 134 non-image variables.
%%
The analysis was carried out for subjects divided by \emph{gender}, \emph{BMI}, and \emph{body size} to assess their influence on the lumbar spine shape.

>>>>

% TODO: BLIND Version
\author{Blind}
% \author{\authorname{Paul Klemm\sup{1}, Sylvia Glaßer\sup{1}, Kai Lawonn\sup{1}, Marko Rak\sup{1} Henry Völzke\sup{2}, Katrin Hegenscheid\sup{3} and Bernhard Preim\sup{1}}
% \affiliation{\sup{1} Department of Simulation and Graphics, University of Magdeburg, Germany}
% \affiliation{\sup{2} Institute for Community Medicine, University of Greifswald, Germany}
% \affiliation{\sup{3} Institute for Diagnostic Radiology and Neuro-Radiology, University of Greifswald, Germany}
% \email{\{klemm, glasser\}@isg.cs.uni-magdeburg.de}
% }

@@

% BLIND
%As described in a prior work (\cite{Klemm2014VIS}), the suggestion of potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
Suggesting potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
%%
We can achieve this also using the presented method by trying to describe several (dichotomous) target values with a decision tree constructed from all available data.
%%
This allows both to assess the discriminative power of the data set regarding the target variables, as well as the most important variables for the classification.
%Except for gender in the age-plot, no variable could be explained with reasonable accuracy for age or body size.
\paragraph{Predictive Power of Image-Derived Variables.}
We showed, that \emph{torsion}, \emph{curvature} and \emph{curvature angle} of the lumbar spine at the presented precision are not enough to characterize lumbar back pain in the SHIP data set.
%%
Our method made it possible to assess their discriminative power, which is largely limited for separating male and female subjects, \emph{nutrition} variables, as well as different disease indicators.
%%
The C4.5 algorithm showed to be an effective tool for evaluating a set of derived metrics regarding their suitability to classify non-image variables.
%%
Over-fitting to the data indicated by complex decision trees have to be taken into account as well.
%%
% \subsection{Proposed Features}
% \label{sec:ProposedFeatures}
% \begin{figure}[!h]
%   %\vspace{-0.2cm}
%   \centering
%   \includegraphics[width=0.475\textwidth]{figures/proposed-features}
%   \caption{
% 	Proposed features.
% 	%%
% 	(a) dented vertebraes are a sign of heavy stress and bone absorption.
% 	%%
% 	(b, c) the spine canal shape can be used to characterize scoliosis (curvature sagittal) and lordosis (curvature coronal) as well as the \emph{Cobb-Angles} \cite{Cobb}.
% 	%%
% 	(c) Spine canal thickness is associated with herniated disks.
% 	%%
% 	}
%   \label{fig:proposedFeatures}
% \end{figure}
% %%
% \noindent Pathological deformation of the lumbar spine is usually the last stage and is associated with very strong lower back pain.
% %%
% We have to capture earlier signs of pathological change in order to derive better predictive models.
% %%
%
% A very early sign of pathological deformation is the \emph{bone resorption in the center of a vertebra}.
% %%
% They change from being block-shaped to be dented in the center (Fig~\ref{fig:proposedFeatures}~a).
% %%
% This information can be obtained by segmenting the whole vertebra or the top and bottom point if each vertebra center.
% %%
% Another valuable variable would be the \emph{spine canal thickness} (Fig~\ref{fig:proposedFeatures}~c).
% %%
% Low spine canal thickness can be an indicator for an impending herniated disk.
% %%
% Both surface texture of the vertebra and thickness of the spine canal are used to diagnose herniated disks.
% %%
% The \emph{overall spine canal shape} is also of interest, since scoliosis and lordosis can be characterized more precise by deriving the Cobb angles from this shape (Fig~\ref{fig:proposedFeatures}~b,~c).
%% ------------------------------------
\paragraph{Future Work.}
%\com{ToDo: Passen die Contributions hier wirklich?}
%% Contributions
We provide a method for comparative analysis of decision trees independent of the variable manifestation count using interactive scatter plots.
%%
We applied the technique to derive insight into the predictive power of 9 image-derived variables for 134 non-image variables.
%%
The analysis was carried out for subjects divided by \emph{gender}, \emph{BMI}, and \emph{body size} to assess their influence on the lumbar spine shape.
